Quasi-universal acceptance of basic science in the US

### 2024-09-19 appendix_exp3.Rmd
submission public understanding of science

Table \@ref(tab:knowledge-overview) shows all questions, their scientifically consensual answer, and their source. All but two questions were selected from existing science knowledge questionnaires. We tried to select non-political questions.

```{r knowledge-overview}
items <- read_csv("exp_3/materials/overview_questions.csv") %>% 
  mutate(id = 1:nrow(.)) %>% 
  select(id, "Question", "Scientific consensus (Study 1)", "Explanation (Study 2 & 3)")

@@

Fig. \@ref(fig:exp3-conditional-acceptance) illustrates the relationship between knowledge and acceptance. In most cases (`r exp3_descriptives$conditional_acceptance$false$acceptance_initial$Yes$share`), participants readily accepted the scientific consensus right after having given the wrong answer to a question. After providing the chance to revise the initial consensus rejection, this share is even larger (`r exp3_descriptives$conditional_acceptance$false$acceptance$Yes$share`). In very few cases (`r exp3_descriptives$conditional_acceptance$true$acceptance_initial$No$share`), participants who initially gave the correct response afterwards rejected the scientific consensus right after, thereby contradicting their own initial response. This share drops slightly after providing the chance to revise the initial consensus rejection (`r exp3_descriptives$conditional_acceptance$true$acceptance$No$share`).

For all correlations, we include opinion revisions for measuring consensus acceptance. We find no statistically significant correlation between science knowledge and trust in science (r = `r exp3_descriptives$cor_trust_knowledge$estimate`, p `r exp3_descriptives$cor_trust_knowledge$p.value`), but a samll positive correlation between acceptance of scientific consensus and trust in science (r = `r exp3_descriptives$cor_trust_acceptance$estimate`, p `r exp3_descriptives$cor_trust_acceptance$p.value`). Again, these findings might be partly due to ceiling effects: As illustrated in Fig. \@ref(fig:exp3-plot), (i) most people do trust science, and (ii) that is true even among people with low knowledge or acceptance rates. We find no statistically significant correlation between conspiracy thinking and science knowledge (r = `r exp3_descriptives$cor_conspiracy_knowledge$estimate`, p `r exp3_descriptives$cor_conspiracy_knowledge$p.value`), and between conspiracy thinking and acceptance of scientific consensus (r = `r exp3_descriptives$cor_conspiracy_acceptance$estimate`, p `r exp3_descriptives$cor_conspiracy_acceptance$p.value`).

@@

All answers can be read in the appendix.

@@

## By-question variation

```{r}
# calculate share
plot_data <- exp3_long %>% 
  group_by(subject) %>% 
  summarize(share_true = sum(knowledge == TRUE, na.rm = TRUE)/n(),
            share_accepted = sum(acceptance == "Yes", na.rm = TRUE)/n()) 
```

### Knowledge

(ref:exp3-questions-knowledge) Distribution of correct answers by question.

```{r exp3-questions-knowledge, fig.cap="(ref:exp3-questions-knowledge)"}
# plot knowledge
ggplot(plot_data, aes(x = reorder(subject, share_true), y = share_true)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format(), breaks = seq(0, 1, 0.1)) +
  labs(x = "Question", y = "Share of correct responses") +
  coord_flip()
```

(ref:exp3-questions-acceptance) Distribution of consensus acceptance by question.

```{r exp3-questions-acceptance, fig.cap="(ref:exp3-questions-acceptance)"}
# plot acceptance
ggplot(plot_data, aes(x = reorder(subject, share_accepted), y = share_accepted)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format(), breaks = seq(0, 1, 0.1)) +
  labs(x = "Question", y = "Share of acceptance") +
  coord_flip()
```

## Trust in science

(ref:exp3-trust-scientists) Distribution of trust in scientists.

```{r exp3-trust-scientists, fig.cap="(ref:exp3-trust-scientists)"}
# calculate share
plot_data <- exp3_wide %>% 
  group_by(wgm_scientists) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  mutate(share = n/sum(n)) %>% 
  rounded_numbers()

# plot knowledge
ggplot(plot_data, aes(x = wgm_scientists, y = share)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format(), breaks = seq(0, 1, 0.1)) +
  labs(x = "Trust in science", y = "Share of participants") +
  geom_text(aes(label = paste0(round(share * 100, 1), "%")),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3) 
```

## Conspiracy thinking

(ref:exp3-conspiracy-distribution) Distribution of conspiracy thinking.

```{r exp3-conspiracy-distribution, fig.cap="(ref:exp3-conspiracy-distribution)"}
# plot knowledge
ggplot(exp3_wide, aes(x = BCTI_avg)) +
  geom_histogram() +
  labs(x = "Average conspiracy thinking (BCTI)", y = "N")
```

### By-item variation

(ref:exp3-conspiracy-items) Distribution of conspiracy thinking by item.

```{r exp3-conspiracy-items, fig.cap="(ref:exp3-conspiracy-items)"}
BCTI_data <- exp3_long %>% 
  # bring to long format
  pivot_longer(cols = BCTI_apollo:BCTI_evolution, 
               names_to = "BCTI_item", values_to = "score") %>% 
  select(id, c(starts_with("BCTI"), score)) %>% 
  group_by(BCTI_item) %>% 
  summarize(avg_score = mean(score, na.rm=TRUE)) %>% 
  ungroup()

ggplot(BCTI_data, aes(x = reorder(BCTI_item, avg_score), y = avg_score)) +
    geom_pointrange(aes(ymin = 1, ymax = avg_score),
                  # Make the lines a little thicker and the dots a little bigger
                  fatten = 2, size = 1) +
  scale_y_continuous(limits = c(1, 9), breaks = seq(1, 9, 1)) +
  labs(x = "Conspiracy item", y = "Average score") +
  coord_cartesian(ylim = c(1, 9)) +
  coord_flip()
```

### All raw answers

# apa_table(table,
#           align = c("m{1cm}", "m{3cm}", "m{10cm}"), 
#           longtable = TRUE
#           )
kbl(table, booktabs = T, longtable = TRUE,
    caption = "Reasons for consensus rejection") %>%
  kable_paper(full_width = F) %>% 
  column_spec(1) %>%
  column_spec(2) %>%
  column_spec(3, width = "30em")

## Why do people accept the scientific consensus:

### Participants who answered "other"

kbl(table, booktabs = T, longtable = TRUE,
    caption = "Reasons for consensus rejection") %>%
  kable_paper(full_width = F) %>% 
  column_spec(1) %>%
  column_spec(2, width = "30em")
```

### Participants who answered "other"

kbl(table, booktabs = T, longtable = TRUE,
    caption = "Reasons for consensus rejection") %>%
  kable_paper(full_width = F) %>% 
  column_spec(1) %>%
  column_spec(2, width = "30em")
```

## Additional plots

(ref:exp3-conditional-acceptance) Acceptance rates of scientific consensus, based on whether the initial response to the knowledge question was false or true.

```{r exp3-conditional-acceptance, fig.cap="(ref:exp3-conditional-acceptance)"}
plot_data <- exp3_long %>% 
  pivot_longer(c(acceptance, acceptance_initial), 
               names_to = "measure", 
               values_to = "acceptance") %>% 
  # give nicer names
  mutate(measure = case_when(measure == "acceptance" ~ "Acceptance including revisions", 
                             measure == "acceptance_initial" ~ "Initial acceptance"), 
         # make a factor
         measure = factor(measure, levels = unique(measure)[c(2, 1)])
         ) %>% 
  group_by(measure, knowledge, acceptance) %>% 
  count() %>% 
  group_by(measure, knowledge) %>% 
  mutate (share = n/sum(n)) %>% 
  mutate_if(is.numeric, round, digits = 3)

absolute_numbers <- plot_data %>% 
  group_by(measure, knowledge) %>% 
  summarise(n = sum(n))

ggplot(plot_data, aes(x = knowledge, y = share, fill = acceptance)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format(), breaks = seq(0, 1, 0.1)) +
  geom_text(aes(label = paste0(round(share * 100, 1), "%")),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3) +
  scale_fill_viridis_d(option = "cividis") +
  labs(x = "Knowledge answer", fill = "Acceptance \nof consensus") +
  plot_theme +
  geom_bracket(data = absolute_numbers,
               aes(xmin = as.numeric(factor(knowledge)) - 0.4, 
                   xmax = as.numeric(factor(knowledge)) + 0.4, 
                   y.position = 1.05, label =  paste0("n = ", n)),
               inherit.aes = FALSE,
               step.increase = 0) +
  facet_wrap(~measure)
```

(ref:exp3-plot-overview) **A** Shows the distribution of science knowledge (left) and acceptance of scientific consensus for participants who gave the wrong answer **B** Shows the relationship between trust in science and science knowledge/acceptance (if wrong at first, rounded to the first digit) of scientific consensus **C** Shows the relationship between conspiracy thinking and science knowledge/acceptance (if wrong at first) of scientific consensus

@@

>>>>

Table \@ref(tab:exp3-knowledge-items) shows all questions, their scientifically consensual answer, and their source. All but two questions were selected from existing science knowledge questionnaires. We tried to select non-political questions.

```{r exp3-knowledge-items}
items <- read_csv("exp_3/materials/overview_questions.csv") %>% 
  mutate(id = 1:nrow(.)) %>% 
  select(id, "Question", "Scientific consensus (Study 1)", "Explanation (Study 2 & 3)")

@@

In most cases (`r exp3_descriptives$conditional_acceptance$false$acceptance_initial$Yes$share`), participants readily accepted the scientific consensus right after having given the wrong answer to a question. After providing the chance to revise the initial consensus rejection, this share is even larger (`r exp3_descriptives$conditional_acceptance$false$acceptance$Yes$share`). In very few cases (`r exp3_descriptives$conditional_acceptance$true$acceptance_initial$No$share`), participants who initially gave the correct response afterwards rejected the scientific consensus right after, thereby contradicting their own initial response. This share drops slightly after providing the chance to revise the initial consensus rejection (`r exp3_descriptives$conditional_acceptance$true$acceptance$No$share`).

For all correlations, we include opinion revisions for measuring consensus acceptance. We find no statistically significant correlation between science knowledge and trust in science (r = `r exp3_descriptives$cor_trust_knowledge$estimate`, p `r exp3_descriptives$cor_trust_knowledge$p.value`), but a samll positive correlation between acceptance of scientific consensus and trust in science (r = `r exp3_descriptives$cor_trust_acceptance$estimate`, p `r exp3_descriptives$cor_trust_acceptance$p.value`). Again, these findings might be partly due to ceiling effects: (i) most people do trust science, and (ii) that is true even among people with low knowledge or acceptance rates. We find no statistically significant correlation between conspiracy thinking and science knowledge (r = `r exp3_descriptives$cor_conspiracy_knowledge$estimate`, p `r exp3_descriptives$cor_conspiracy_knowledge$p.value`), and between conspiracy thinking and acceptance of scientific consensus (r = `r exp3_descriptives$cor_conspiracy_acceptance$estimate`, p `r exp3_descriptives$cor_conspiracy_acceptance$p.value`).

@@

All answers can be read in the below.

@@

(ref:exp3-plot-overview) Summary plot for Study 3. **A** Shows the distribution of science knowledge (left) and acceptance of scientific consensus for participants who gave the wrong answer **B** Shows the relationship between trust in science and science knowledge/acceptance (if wrong at first, rounded to the first digit) of scientific consensus **C** Shows the relationship between conspiracy thinking and science knowledge/acceptance (if wrong at first) of scientific consensus

@@

# apa_table(table,
#           align = c("m{1cm}", "m{3cm}", "m{10cm}"), 
#           longtable = TRUE
#           )
kbl(table, booktabs = T, longtable = TRUE,
    caption = "Reasons for consensus rejection (raw answers)") %>%
  kable_paper(full_width = F) %>% 
  column_spec(1) %>%
  column_spec(2) %>%
  column_spec(3, width = "30em")

## Reasons for consensus acceptance

kbl(table, booktabs = T, longtable = TRUE,
    caption = "Reasons for consensus acceptance (participants who answered 'other')") %>%
  kable_paper(full_width = F) %>% 
  column_spec(1) %>%
  column_spec(2, width = "30em")
```

kbl(table, booktabs = T, longtable = TRUE,
    caption = "Answers from participants saying they trusted the scientific consens because they verified independently on how they did so.") %>%
  kable_paper(full_width = F) %>% 
  column_spec(1) %>%
  column_spec(2, width = "30em")
```

### 2024-09-19 appendix_exp4.Rmd
submission public understanding of science

Fig. \@ref(fig:exp4-conditional-acceptance) illustrates the relationship between knowledge and acceptance. In most cases (`r exp4_descriptives$conditional_acceptance$false$acceptance_initial$Yes$share`), participants readily accepted the scientific consensus right after having given the wrong answer to a question. After providing the chance to revise the initial consensus rejection, this share is even larger (`r exp4_descriptives$conditional_acceptance$false$acceptance$Yes$share`). In few (but still perhaps surprisingly many) cases (`r exp4_descriptives$conditional_acceptance$true$acceptance_initial$No$share`), participants who initially gave the correct response afterwards rejected the scientific consensus right after, thereby contradicting their own initial response. This share drops slightly after providing the chance to revise the initial consensus rejection (`r exp4_descriptives$conditional_acceptance$true$acceptance$No$share`).

For all correlations, we include opinion revisions for measuring consensus acceptance. We find no statistically significant correlation between science knowledge and trust in science (r = `r exp4_descriptives$cor_trust_knowledge$estimate`, p `r exp4_descriptives$cor_trust_knowledge$p.value`), but a small positive correlation between acceptance of scientific consensus and trust in science (r = `r exp4_descriptives$cor_trust_acceptance$estimate`, p `r exp4_descriptives$cor_trust_acceptance$p.value`). Again, these findings might be partly due to ceiling effects: As illustrated in Fig. \@ref(fig:exp4-plot), (i) most people do trust science, and (ii) that is true even among people with low knowledge or acceptance rates. We do not find a statistically significant correlation between neither conspiracy thinking and science knowledge (r = `r exp4_descriptives$cor_conspiracy_knowledge$estimate`, p `r exp4_descriptives$cor_conspiracy_knowledge$p.value`), nor between conspiracy thinking and acceptance of scientific consensus (r = `r exp4_descriptives$cor_conspiracy_acceptance$estimate`, p `r exp4_descriptives$cor_conspiracy_acceptance$p.value`).

@@

## By-question variation

```{r}
# calculate share
plot_data <- exp4_long %>% 
  group_by(subject) %>% 
  summarize(share_true = sum(knowledge == TRUE, na.rm = TRUE)/n(),
            share_accepted = sum(acceptance == "Yes", na.rm = TRUE)/n()) 
```

### Knowledge

(ref:exp4-questions-knowledge) Distribution of correct answers by question.

```{r exp4-questions-knowledge, fig.cap="(ref:exp4-questions-knowledge)"}
# plot knowledge
ggplot(plot_data, aes(x = reorder(subject, share_true), y = share_true)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format(), breaks = seq(0, 1, 0.1)) +
  labs(x = "Question", y = "Share of correct responses") +
  coord_flip()
```

(ref:exp4-questions-acceptance) Distribution of consensus acceptance by question.

```{r exp4-questions-acceptance, fig.cap="(ref:exp4-questions-acceptance)"}
# plot acceptance
ggplot(plot_data, aes(x = reorder(subject, share_accepted), y = share_accepted)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format(), breaks = seq(0, 1, 0.1)) +
  labs(x = "Question", y = "Share of acceptance") +
  coord_flip()
```

## Trust in science

(ref:exp4-trust-scientists) Distribution of trust in scientists.

```{r exp4-trust-scientists, fig.cap="(ref:exp4-trust-scientists)"}
# calculate share
plot_data <- exp4_wide %>% 
  group_by(wgm_scientists) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  mutate(share = n/sum(n)) %>% 
  rounded_numbers()

# plot knowledge
ggplot(plot_data, aes(x = wgm_scientists, y = share)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format(), breaks = seq(0, 1, 0.1)) +
  labs(x = "Trust in science", y = "Share of participants") +
  geom_text(aes(label = paste0(round(share * 100, 1), "%")),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3) 
```

## Conspiracy thinking

(ref:exp4-conspiracy-distribution) Distribution of conspiracy thinking.

```{r exp4-conspiracy-distribution, fig.cap="(ref:exp4-conspiracy-distribution)"}
# plot knowledge
ggplot(exp4_wide, aes(x = BCTI_avg)) +
  geom_histogram() +
  labs(x = "Average conspiracy thinking (BCTI)", y = "N")
```

### By-item variation

(ref:exp4-conspiracy-items) Distribution of conspiracy thinking by item.

```{r exp4-conspiracy-items, fig.cap="(ref:exp4-conspiracy-items)"}
BCTI_data <- exp4_long %>% 
  # bring to long format
  pivot_longer(cols = BCTI_apollo:BCTI_evolution, 
               names_to = "BCTI_item", values_to = "score") %>% 
  select(id, c(starts_with("BCTI"), score)) %>% 
  group_by(BCTI_item) %>% 
  summarize(avg_score = mean(score, na.rm=TRUE)) %>% 
  ungroup()

ggplot(BCTI_data, aes(x = reorder(BCTI_item, avg_score), y = avg_score)) +
    geom_pointrange(aes(ymin = 1, ymax = avg_score),
                  # Make the lines a little thicker and the dots a little bigger
                  fatten = 2, size = 1) +
  scale_y_continuous(limits = c(1, 9), breaks = seq(1, 9, 1)) +
  labs(x = "Conspiracy item", y = "Average score") +
  coord_cartesian(ylim = c(1, 9)) +
  coord_flip()
```

### All raw answers

# apa_table(table,
#           align = c("m{1cm}", "m{3cm}", "m{10cm}"), 
#           longtable = TRUE
#           )
kbl(table, booktabs = T, longtable = TRUE,
    caption = "Reasons for consensus rejection") %>%
  kable_paper(full_width = F) %>% 
  column_spec(1) %>%
  column_spec(2) %>%
  column_spec(3, width = "30em")

## Why do people accept the scientific consensus:

### Participants who answered "other"

kbl(table, booktabs = T, longtable = TRUE,
    caption = "Reasons for consensus rejection") %>%
  kable_paper(full_width = F) %>% 
  column_spec(1) %>%
  column_spec(2, width = "30em")
```

### Participants who answered "other"

kbl(table, booktabs = T, longtable = TRUE,
    caption = "Reasons for consensus rejection") %>%
  kable_paper(full_width = F) %>% 
  column_spec(1) %>%
  column_spec(2, width = "30em")
```

## Additional plots

(ref:exp4-conditional-acceptance) Acceptance rates of scientific consensus, based on whether the initial response to the knowledge question was false or true.

```{r exp4-conditional-acceptance, fig.cap="(ref:exp4-conditional-acceptance)"}
plot_data <- exp4_long %>% 
  pivot_longer(c(acceptance, acceptance_initial), 
               names_to = "measure", 
               values_to = "acceptance") %>% 
  # give nicer names
  mutate(measure = case_when(measure == "acceptance" ~ "Acceptance including revisions", 
                             measure == "acceptance_initial" ~ "Initial acceptance"), 
         # make a factor
         measure = factor(measure, levels = unique(measure)[c(2, 1)])
         ) %>% 
  group_by(measure, knowledge, acceptance) %>% 
  count() %>% 
  group_by(measure, knowledge) %>% 
  mutate (share = n/sum(n)) %>% 
  mutate_if(is.numeric, round, digits = 3)

absolute_numbers <- plot_data %>% 
  group_by(measure, knowledge) %>% 
  summarise(n = sum(n))

ggplot(plot_data, aes(x = knowledge, y = share, fill = acceptance)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format(), breaks = seq(0, 1, 0.1)) +
  geom_text(aes(label = paste0(round(share * 100, 1), "%")),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3) +
  scale_fill_viridis_d(option = "cividis") +
  labs(x = "Knowledge answer", fill = "Acceptance \nof consensus") +
  plot_theme +
  geom_bracket(data = absolute_numbers,
               aes(xmin = as.numeric(factor(knowledge)) - 0.4, 
                   xmax = as.numeric(factor(knowledge)) + 0.4, 
                   y.position = 1.05, label =  paste0("n = ", n)),
               inherit.aes = FALSE,
               step.increase = 0) +
  facet_wrap(~measure)
```

(ref:exp4-plot-overview) **A** Shows the distribution of science knowledge (left) and acceptance of scientific consensus for participants who gave the wrong answer **B** Shows the relationship between trust in science and science knowledge/acceptance (if wrong at first, rounded to the first digit) of scientific consensus **C** Shows the relationship between conspiracy thinking and science knowledge/acceptance (if wrong at first) of scientific consensus

>>>>

In most cases (`r exp4_descriptives$conditional_acceptance$false$acceptance_initial$Yes$share`), participants readily accepted the scientific consensus right after having given the wrong answer to a question. After providing the chance to revise the initial consensus rejection, this share is even larger (`r exp4_descriptives$conditional_acceptance$false$acceptance$Yes$share`). In few (but still perhaps surprisingly many) cases (`r exp4_descriptives$conditional_acceptance$true$acceptance_initial$No$share`), participants who initially gave the correct response afterwards rejected the scientific consensus right after, thereby contradicting their own initial response. This share drops slightly after providing the chance to revise the initial consensus rejection (`r exp4_descriptives$conditional_acceptance$true$acceptance$No$share`).

For all correlations, we include opinion revisions for measuring consensus acceptance. We find no statistically significant correlation between science knowledge and trust in science (r = `r exp4_descriptives$cor_trust_knowledge$estimate`, p `r exp4_descriptives$cor_trust_knowledge$p.value`), but a small positive correlation between acceptance of scientific consensus and trust in science (r = `r exp4_descriptives$cor_trust_acceptance$estimate`, p `r exp4_descriptives$cor_trust_acceptance$p.value`). Again, these findings might be partly due to ceiling effects: (i) most people do trust science, and (ii) that is true even among people with low knowledge or acceptance rates. We do not find a statistically significant correlation between neither conspiracy thinking and science knowledge (r = `r exp4_descriptives$cor_conspiracy_knowledge$estimate`, p `r exp4_descriptives$cor_conspiracy_knowledge$p.value`), nor between conspiracy thinking and acceptance of scientific consensus (r = `r exp4_descriptives$cor_conspiracy_acceptance$estimate`, p `r exp4_descriptives$cor_conspiracy_acceptance$p.value`).

@@

(ref:exp4-plot-overview) Summary plot for Study 4. **A** Shows the distribution of science knowledge (left) and acceptance of scientific consensus for participants who gave the wrong answer **B** Shows the relationship between trust in science and science knowledge/acceptance (if wrong at first, rounded to the first digit) of scientific consensus **C** Shows the relationship between conspiracy thinking and science knowledge/acceptance (if wrong at first) of scientific consensus
